Regression models for Cylindrical data in Psychology

### 2019-03-06 Manuscript/Manuscript.tex
Manuscript Update

\linenumbers
\DeclareRobustCommand{\VANDER}[3]{#2}
\DeclareRobustCommand{\VAN}[3]{#2}
\DeclareRobustCommand{\DEN}[3]{#2}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{color}
\usepackage{subcaption}

@@

\section{Introduction}\label{Introduction}

In social sciences the use of cylindrical data is very common. Such data consist
of a linear and a circular component. Gurtman (2011) refers to such data
as vectors, with a directional measure (i.e., the circular component) and a
measure indicating the magnitude (i.e., the linear component). Many established
models in psychology are often referred to as circular or circumplex models, but
those models are cylindrical. Examples of such cylindrical models are the
interpersonal circle/circumplex
(Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans, \DEN{Brok}{Den}{den} Brok, \& \VAN{Tartwijk}{Van}{van} Tartwijk, 2006), the circumplex of
affect (Russell, 1980), the circumplex of human emotion
(Plutchik, 1997) or the model of human values (Schwartz, 1992).
\newline
\indent Also, many of the more recent types of data that are studied in psychology are
cylindrical. For example, research on human navigation uses data where distance
(i.e., linear component) and direction (i.e., the circular component) are of
interest (Chrastil \& Warren, 2017) or in eye-tracking, the saccade data also
consist of both the direction (i.e., the circular variable) and the duration
(i.e., the linear variable) (e.g., Rayner (2009)). Apart from the social
sciences, data with a circular and linear outcome more commonly occur in
meteorology (García-Portugués, Crujeiras, \& González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras, González-Manteiga, \& Pereira, 2014) or
marine research (Lagona, Picone, Maruotti, \& Cosoli, 2015)
\newline
\indent Up until now researchers studying cylindrical data had to rely on linear
statistical methods to analyze their research results. However, lately more and more
of these researchers acknowledge that linear methods are not sufficient and call
for new methods (Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy, \& Hilsenroth, 2009) that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill
the above mentioned gap in the literature by showing that the use of cylindrical
models can benefit the analysis of circumplex data and cylindrical data in
psychology in general. More specifically we will show these benefits for
interpersonal teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components from a set
of covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto a set
of covariates. From now on we will therefore refer to the components of the
cylindrical variable as outcome components. These adapted cylindrical models are
then used to analyse the teacher data.

\subsection{Cylindrical data}

@@

\subsection{Data Description}\label{DataDescriptives}

@@

\section{Four Cylindrical Regression Models}\label{Models}

@@

\subsection{The Modified Circular-Linear Projected Normal (CL-PN) and Modified Circular-Linear General Projected Normal (CL-GPN) Models}\label{CL-(G)PN}

@@

\subsection{The Modified Abe-Ley Model}\label{WeiSSVM}

@@

\subsection{Modified Joint Projected and Skew Normal Model (GPN-SSN)}\label{CL-GPN_multivariate}

@@

\noindent The GPN-SSN model thus allows us to assess the average type and
strength of interpersonal behavior through the parameters \(\beta_{0_{s^{I}}}\),
\(\beta_{0_{s^{II}}}\) and \(\beta_{0_{y}}\) as well as the effect of self-efficacy
on type and strength of teacher behavior through the parameters
\(\beta_{1_{s^{I}}}\), \(\beta_{1_{s^{II}}}\) and \(\beta_{1_{y}}\). In addition,
because the type and strength of interpersonal behavior are modelled together
using a multivariate normal distribution we can, through its variance-covariance
matrix, also assess the dependence between the type and strength of interpersonal
behavior. \newline
\indent We estimate the model using MCMC methods. A detailed description of these
methods is given in the Supplementary Material.

\subsection{Model Fit Criterion}\label{Modelfit}

@@

\section{Data Analysis}\label{DataAnalysis}

@@

\subsection{Results \& Analysis}\label{DataResults}

@@

\subsubsection{Average type and strength of interpersonal behavior}

@@

\subsubsection{The effect of self-efficacy}

@@

score of the teacher on the IPC moves counterclockwise.\newline

\begin{figure}
\centering
\includegraphics[width = \textwidth]{Plots/reglinediffSE.pdf}
\caption{Plot showing circular regression lines for the effect of self-efficacy
as predicted by the Abe-Ley model (solid line) and CL-PN model (dashed line). The
black square indicates the inflection point of the circular regression line for
the CL-PN model.}
\label{regline}
\end{figure}

\subsubsection{Dependence between type and strength of interpersonal behavior.}

@@

\subsubsection{Model fit}

@@

\section{Discussion}\label{Discussion}

@@

developed in the present study.\newline

\section*{References}

>>>>

\linenumbers
\DeclareRobustCommand{\VANDER}[3]{#2}
\DeclareRobustCommand{\VAN}[3]{#2}
\DeclareRobustCommand{\DEN}[3]{#2}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{color}
\usepackage{subcaption}
\raggedbottom

@@

\vspace{-0.75cm}

In social sciences the use of cylindrical data is very common. Such data consist
of a linear and a circular component. Gurtman (2011) refers to such data
as vectors, with a directional measure (i.e., the circular component) and a
measure indicating the magnitude (i.e., the linear component). Many established
models in psychology are often referred to as circular or circumplex models, but
those models are cylindrical. Examples of such cylindrical models are the
interpersonal circle/circumplex
(Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans, \DEN{Brok}{Den}{den} Brok, \& \VAN{Tartwijk}{Van}{van} Tartwijk, 2006), the circumplex of
affect (Russell, 1980), the circumplex of human emotion
(Plutchik, 1997) or the model of human values (Schwartz, 1992).
\newline
\indent Also, many of the more recent types of data that are studied in psychology are
cylindrical. For example, research on human navigation uses data where distance
(i.e., the linear component) and direction (i.e., the circular component) are of
interest (Chrastil \& Warren, 2017) or in eye-tracking, the saccade data also
consist of both the direction (i.e., the circular variable) and the duration
(i.e., the linear variable) (e.g., Rayner (2009)). Apart from the social
sciences, data with a circular and linear outcome more commonly occur in
meteorology (García-Portugués, Crujeiras, \& González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras, González-Manteiga, \& Pereira, 2014) or
marine research (Lagona, Picone, Maruotti, \& Cosoli, 2015)
\newline
\indent Up until now researchers studying cylindrical data had to rely on linear
statistical methods to analyze their research results. However, lately more and more
of these researchers acknowledge that linear methods are not sufficient and call
for new methods (Gurtman, 2011; Pennings, 2017b; Wright, Pincus, Conroy, \& Hilsenroth, 2009) that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill
the above mentioned gap in the literature by showing that the use of cylindrical
models can benefit the analysis of circumplex data and cylindrical data in
psychology in general. More specifically we will show these benefits for
interpersonal teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components from a set
of covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto a set
of covariates. From now on we will therefore refer to the components of the
cylindrical variable as outcome components. These adapted cylindrical models are
then used to analyse the teacher data.

\vspace{-0.75cm}
\subsection{Cylindrical Data}
\vspace{-0.75cm}

@@

\vspace{-0.75cm}
\subsection{Data Description}\label{DataDescriptives}
\vspace{-0.75cm}

@@

\vspace{-0.75cm}
\section{Four Cylindrical Regression Models}\label{Models}
\vspace{-0.75cm}

@@

\vspace{-0.75cm}
\subsection{The Modified Circular-Linear Projected Normal (CL-PN) and Modified Circular-Linear General Projected Normal (CL-GPN) Models}\label{CL-(G)PN}
\vspace{-0.75cm}

@@

\vspace{-0.75cm}
\subsection{The Modified Abe-Ley Model}\label{WeiSSVM}
\vspace{-0.75cm}

@@

\vspace{-0.75cm}
\subsection{Modified Joint Projected and Skew Normal Model (GPN-SSN)}\label{CL-GPN_multivariate}
\vspace{-0.75cm}

@@

\noindent The GPN-SSN model thus allows us to assess the average type and
strength of interpersonal behavior through the parameters \(\beta_{0_{s^{I}}}\),
\(\beta_{0_{s^{II}}}\) and \(\beta_{0_{y}}\) as well as the effect of self-efficacy
on type and strength of teacher behavior through the parameters
\(\beta_{1_{s^{I}}}\), \(\beta_{1_{s^{II}}}\) and \(\beta_{1_{y}}\). In addition,
because the type and strength of interpersonal behavior are modelled together
using a multivariate normal distribution we can, through its variance-covariance
matrix, also assess the dependence between the type and strength of interpersonal
behavior. \newpage
\indent We estimate the model using MCMC methods. A detailed description of these
methods is given in the Supplementary Material.

\vspace{-0.75cm}
\subsection{Model Fit Criterion}\label{Modelfit}
\vspace{-0.75cm}

@@

\vspace{-0.75cm}
\section{Results}\label{DataAnalysis}
\vspace{-0.75cm}

@@

\vspace{-0.75cm}
\subsection{Analysis}\label{DataResults}
\vspace{-0.75cm}

@@

\vspace{-0.5cm}
\subsubsection{Average type and strength of interpersonal behavior}

@@

\vspace{-0.5cm}
\subsubsection{The effect of self-efficacy}

@@

score of the teacher on the IPC moves counterclockwise.

\begin{figure}
\centering
\includegraphics[width = \textwidth]{Plots/reglinediffSE.pdf}
\caption{Plot showing circular regression lines for the effect of self-efficacy
as predicted by the Abe-Ley model (solid line) and CL-PN model (dashed line). The
black square indicates the inflection point of the circular regression line for
the CL-PN model.}
\label{regline}
\end{figure}
\vspace{-0.5cm}
\subsubsection{Dependence between type and strength of interpersonal behavior.}

@@

\vspace{-0.5cm}
\subsubsection{Model fit}

@@

\vspace{-0.75cm}
\section{Discussion}\label{Discussion}
\vspace{-0.75cm}

@@

developed in the present study.

\newpage
\vspace{-0.75cm}
\section*{{\normalfont References}}
\vspace{-0.75cm}
